---
title: "Pan-human Azimuth in R"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: united
    df_print: kable
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
---

```{r setup, include=FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))

knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  fig.align = "center",
  message = FALSE,
  warning = FALSE,
  time_it = TRUE,
  error = TRUE
)
```

This vignette demonstrates how to install and use `AzimuthAPI`, an interface for cell type annotation using the Pan-human Azimuth neural network.

The API takes a Seurat object and returns hierarchical cell type predictions, confidence scores, and low-dimensional embeddings useful for visualization.

# Install AzimuthAPI

The Pan-human Azimuth R API, `AzimuthAPI`, is available for installation via Github:

```{r install.panazimuth, eval=FALSE}
remotes::install_github("satijalab/AzimuthAPI")
```

# Load packages & data

```{r load.packages}
library(AzimuthAPI)
library(Seurat)
library(SeuratData)
library(ggplot2)
library(dplyr)
```

For this vignette, we will demonstrate annotating a dataset of human bone marrow mononuclear (BMNC) cells that we published as part of [Stuart, Butler et al Cell 2019](https://doi.org/10.1016/j.cell.2019.05.031). This dataset is available through `SeuratData`.

```{r install.data, eval = FALSE}
bmcite <- InstallData("bmcite")
```

Load the data from `SeuratData`. Here we also make sure to normalize the data before running Pan-human Azimuth.

```{r load.data}
bmcite <- LoadData("bmcite")
bmcite <- NormalizeData(bmcite)
```

# Annotate a dataset with Pan-human Azimuth

Users have two options for running Pan-human Azimuth on their dataset: cloud-based or local. The cloud-based option is the easiest to use, and requires no additional setup. The local option requires setting up a Python environment with panhumanpy (Python package for the Pan-human Azimuth neural network) and its dependencies.

Annotation results, regardless of the method used, are stored in the Seurat object as cell-level metadata, and the embeddings generated by the underlying neural network are stored in a new Seurat reduction.

## Option 1: run `CloudAzimuth`

The `CloudAzimuth` function runs Pan-human Azimuth cell type annotation on a Seurat object via a cloud-based API deployed on AWS.

```{r run.panazimuth, results = 'hide', message = FALSE, warning = FALSE}
bmcite <- CloudAzimuth(bmcite)
```

## Option 2: run `ANNotate`

First, set up a local Python environment:

```{bash setup.local.env, eval = FALSE}
# Create a new conda environment for Pan-human Azimuth
conda create -n pan-human-azimuth python=3.9
# Activate the environment
conda activate pan-human-azimuth
# Install Pan-human Azimuth and its dependencies
# for more installation options, see https://satijalab.org/pan_human_azimuth/python/panhumanpy.html
pip install panhumanpy
```

Then ask `reticulate` to use that environment:

```{r use.local.env}
library(reticulate)
# Use the local Python environment for Pan-human Azimuth
use_condaenv("pan-human-azimuth", required = TRUE)
```

Finally, run the `ANNotate` function to annotate your dataset. Note that the initial run may take longer, as the model weights are downloaded and cached locally.

```{r run.local.panazimuth, results = 'hide', message = FALSE, warning = FALSE}
bmcite <- ANNotate(bmcite)
```

# Examine model output

Pan-human Azimuth returns predictions in multiple formats, as well as softmax probability scores to estimate model confidence.

For more information on the output, see the reference.

```{r}
# Inspect cell-level metadata returned by the model
md <- bmcite[[]]
# New columns - ex. final_level_labels, final_level_confidence, full_hierarchical_labels - have been added to the Seurat object metadata
colnames(md)
```

We can view a histogram of softmax probabilities, which reflect the model confidence associated with each cell's annotation:

```{r softmax.histogram}
ggplot(md, aes(x = final_level_confidence)) + 
    geom_histogram(bins = 20, fill = "skyblue") + 
    labs(x = "Softmax probability", y = "Count", title = "Histogram of Softmax Probabilities") +
    theme_bw()
```

As a QC / filtering step, we can remove cells with low scores:

```{r subset.low.scores}
# remove cells with low scores
bmcite_qc <- subset(bmcite, final_level_confidence > 0.5)
```

# Generate UMAP of Azimuth embeddings

Pan-human Azimuth provides embeddings of the encoding layer generated by the underlying neural network, stored in the `azimuth_embed` 128-dimensional reduction. We can use this to generate a two-dimensional visualization of the dataset, to examine cell type predictions in this low-dimensional space.

```{r run.umap}
# Use azimuth_embed as input to generate UMAP
bmcite_qc <- RunUMAP(bmcite_qc, dims = 1:128, reduction = "azimuth_embed", reduction.name = "azimuth_umap")
```

# Visualize predictions

The full_hierarchical_label for each cell provides the model's classification at each level of granularity, with different levels separated by the '|' character.

```{r pha.labels.1, fig.width = 6, fig.height = 6}
p2 <- DimPlot(bmcite_qc, group.by = "full_hierarchical_labels", label.size = 1.5, label = T, reduction = "azimuth_umap",
    repel = TRUE) + NoLegend()
p2
```

To avoid the long length of the full hierarchical label, we also output the final level of granularity in a separate column, `final_level_labels`, which is the last (most specific)level of the hierarchy for each cell.

```{r pha.labels.2, fig.width = 6, fig.height = 6}
p3 <- DimPlot(bmcite_qc, group.by = "final_level_labels", label.size = 2, label = T, reduction = "azimuth_umap") +
    NoLegend()
p3
```

```{r pha.confidence, fig.width = 6, fig.height = 6}
p4 <- FeaturePlot(bmcite_qc, features = "final_level_confidence", reduction = "azimuth_umap")
p4
```

We also postprocess our predictions to provide labels at three consistent levels of granularity for easy handling, marking any cell with an invalid full hierarchical label (based on `full_consistent_hierarchy`) as `False`.

- `azimuth_broad`: Corresponds to level_zero_labels (e.g. Immune cell)
- `azimuth_medium`: Medium level of granularity (e.g. T cell)
- `azimuth_fine`: High level of granularity (e.g. Treg cell)

These categories provide a consistent level of granularity for each cell, but may differ from the `final_level_label`, either by forcing the model to predict further along the cell type hierarchy than its intial prediction, or by rolling back its prediction to a lower level of granularity.

```{r refined.labels.medium, fig.width = 6, fig.height = 6}
p5 <- DimPlot(bmcite_qc, group.by = "azimuth_medium", label.size = 3, label = T, reduction = "azimuth_umap") +
    NoLegend()
p5
```

```{r refined.labels.fine, fig.width = 6, fig.height = 6}
p6 <- DimPlot(bmcite_qc, group.by = "azimuth_fine", label.size = 3, label = T, reduction = "azimuth_umap") +
    NoLegend()
p6
```

To remove the number of labels displayed, we can filter labels with less than a certain number of cells using `PrepLabel`. 

Here, we filter labels with less than 20 cells per label. This can be useful to filter outliers, especially as Pan-human Azimuth does not perform smoothing of single-cell labels by cluster. Therefore, a single outlier annotation for one cell will still display the outlier label on a visualization, and the `PrepLabel` function can help with this.

```{r prepped.labels, fig.width = 6, fig.height = 6}
bmcite_qc <- PrepLabel(bmcite_qc, "azimuth_fine", "azimuth_fine_filtered", cutoff = 20)
p7 <- DimPlot(bmcite_qc, group.by = "azimuth_fine_filtered", label.size = 3, label = T, reduction = "azimuth_umap") +
    NoLegend()
p7
```

# Visualize differentially expressed features

The `make_azimuth_QC_heatmaps` function allows you to easily explore the quality of predicted labels by creating expression heatmaps by predicted cell type, with optional parameters for improved visualization.

Plots are saved by `azimuth_broad` categories by default, with the exception of immune cell types grouped separately by lymphoid or myeloid/erythroid subpopulations.

```{r make.heatmaps}
plots <- make_azimuth_QC_heatmaps(bmcite)
print(length(plots))
```

```{r lymphoid.heatmap, fig.height = 12, fig.width = 12}
p8 <- plots[['Immune_Lymphoid cell_1']]
print(p8)
```

```{r myeloid.heatmap, fig.height = 12, fig.width = 12}
p9 <- plots[['Immune_Myeloid cell_1']]
print(p9)
```

# More information

- For detailed options and usage of the `AzimuthAPI` functions, see the AzimuthAPI [function reference](https://google.com).
- For another example of the usage of this package, see the [Analysis and visualization of Visium HD data with cell segmentations](https://satijalab.org/seurat/articles/visiumhd_analysis_cell_segmentations) vignette on the Seurat website.

<details> <summary> **Session Info** </summary>

```{r session.info}
sessionInfo()
```

</details>