---
title: "Pan-human Azimuth R API Vignette"
output:
  html_document:
    theme: united
    df_print: kable
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
---
```{r setup, include=FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  fig.align = "center",
  message = TRUE,
  warning = TRUE,
  time_it = TRUE,
  error = TRUE
)
```

# Setup the Environment
Install the Pan-human Azimuth R API from Github
```{r install.panazimuth, eval=FALSE}
devtools::install_github("satijalab/AzimuthAPI", auth_token = "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")
```

# Run ANNotate
For this vignette, we will demonstrate annotating a dataset of human bone marrow mononuclear (BMNC) cells that we published as part of (Stuart, Butler et al Cell 2019)[https://www.cell.com/cell/fulltext/S0092-8674(19)30559-8]. This dataset is available through SeuratData.
```{r load.packages}
library(Seurat)
library(SeuratData)
library(AzimuthAPI)
library(dplyr)
```
```{r install.data, eval = FALSE}
bmcite <- InstallData("bmcite")
```

```{r load.data}
bmcite <- LoadData("bmcite")
```

The `CloudAzimuth` function runs Pan-human Azimuth cell type predictions on a Seurat object via a cloud-based API deployed on AWS, and returns results stored in the object's cell-level metadata. 

```{r run.panazimuth, results = FALSE, cache=TRUE}
bmcite <- CloudAzimuth(bmcite)
```

# Visualize predictions
You can visualize the predicted cell type labels on `azimuthumap`, the UMAP reduction of the embedding that the PanAzimuth model learns on the data based on cell type representation.

For example, below are the original cell type annotations visualized on `azimuthumap`:

```{r orig.labels, fig.width = 9}
p0 <- DimPlot(bmcite, group.by = "celltype.l1", label.size = 3, repel=T, label=T) + NoLegend()
p1 <- DimPlot(bmcite, group.by = "celltype.l2", label.size = 3, repel=T, label=T) + NoLegend()
p0 | p1
```

Pan-human Azimuth returns predictions in multiple formats, as well as softmax probability scores to estimate model confidence:

* `full_hierarchical_labels`: The full predicted label for each cell, where '|' delimits each hierarchical level
* `level_zero_labels`: The most broad level of the full hierarchical label
* `final_level_labels`: The most granular level of the full hierarchical label
* `final_level_softmax_prob`: The model’s predicted probability for each cell’s assigned `final_level_label`, ranging from 0 to 1

In addition, we include the `full_consistent_hierarchy` flag to indicate whether if the full hierarchical label is valid according to our internal cell type hierarchy.

```{r}
# All cell-level metadata that is returned
# Note that the first 10 columns existed in the object prior to running Azimuth
colnames(bmcite[[]])
```
```{r, include = FALSE}
bmcite <- PrepLabel(bmcite, "full_hierarchical_labels", "full_hierarchical_labels", cutoff=20)
bmcite <- PrepLabel(bmcite, "final_level_labels", "final_level_labels", cutoff=20)
```

The `full_hierarchical_label` for each cell provides highly detailed information, but may be difficult to interpret visually compared to level-wise labels. 

```{r panazimuth.labels.1}
p2 <- DimPlot(bmcite, group.by = "full_hierarchical_labels", label.size = 1.5, repel=T, label=T) + NoLegend()
p2
```
```{r panazimuth.labels.2, fig.width = 9}
p3 <- DimPlot(bmcite, group.by = "final_level_labels", label.size = 3, repel=T, label=T) + NoLegend()
p4 <- FeaturePlot(bmcite, features = "final_level_softmax_prob") 
p3 | p4
```

We also postprocess our predictions to provide labels at three consistent levels of granularity for easy handling, marking any cell with an invalid full hierarchical label (based on `full_consistent_hierarchy`) as `False`.  

* `azimuth_broad`: Corresponds to level_zero_labels (i.e. Immune cell)
* `azimuth_medium`: Medium level of granularity (i.e. T cell)
* `azimuth_fine`: High level of granularity (i.e. Memory CD4 T cell)

For `azimuth_medium` and `azimuth_fine`, labels have been adjusted to return labels at different levels in the hierarchy to be consistent with common understandings of cell type granularity. 
This means that for a certain number of cells, an `azimuth_medium` and/or `azimuth_fine` label could be more granular than its initial prediction. In such cases, these labels are likely to be an accurate description of the cell, as we use internal calculations based on probability values returned by the model, but should be handled with caution for more in-depth or formal analysis. 

```{r refined.labels.medium}
p5 <- DimPlot(bmcite, group.by = "azimuth_medium", label.size = 3, repel=T, label=T) + NoLegend()
p5
```
```{r refined.labels.fine}
p6 <- DimPlot(bmcite, group.by = "azimuth_fine", label.size = 3, repel=T, label=T) + NoLegend()
p6
```

To remove the number of labels displayed, you can filter labels with less than a certain number of cells using `PrepLabel`.

```{r prepped.labels}
bmcite <- PrepLabel(bmcite, "azimuth_fine", "azimuth_fine_filtered", cutoff=20)
p7 <- DimPlot(bmcite, group.by = "azimuth_fine_filtered", label.size = 3, repel=T, label=T) + NoLegend()
p7
```

# Visualize differentially expressed features
The `make_azimuth_QC_heatmaps` function allows you to easily explore the quality of predicted labels by creating expression heatmaps by predicted cell type, with optional parameters for improved visualization:

* `final_name`: Name of metadata columns to group cells by (default is `azimuth_fine`)
* `min.final.group`: Minimum number of cells under a cell type to be displayed
* `max.ids.per.plot`: Number of cell type labels displayed per plot
* `reorder`: Flag to indicate whether to reorder cell types by transcriptional similarity
* `cells.order`: Cell names to specify order of cells (i.e. by softmax probability)
* `save_folder_path`: Save plots as PNG files under specified folder path

Plots are saved by `azimuth_broad` categories by default, with the exception of immune cell types grouped separately by lymphoid or myeloid/erythroid subpopulations.

```{r make.heatmaps}
plots <- make_azimuth_QC_heatmaps(bmcite)
print(length(plots))
```
```{r lymphoid.heatmap, fig.height = 24}
p8 <- plots[['Immune_Lymphoid cell_1']]
print(p8)
```
```{r myeloid.heatmap, fig.height = 24}
p9 <- plots[['Immune_Myeloid cell_1']]
print(p9)
```